---
title: "Re-analysis of Lombard et al. 2016"
author: "Luiz Felipe Martucci"
date: "2/16/2022"
output:  
  html_document:
       keep_md: true
bibliography: references.bib
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data analysis of: Preventing Weight Gain in Women in Rural Communities: A Cluster Randomised Controlled Trial. Lombard et al. (2016) [@lombard2016]

Data available at: <https://figshare.com/ndownloader/files/2600235>

```{r libraries, include=FALSE}
#Libraries
x <- c("tidyverse", "foreign", "kableExtra", "Hmisc", "broom", "reshape2")
(function(x){
  sapply(x, function(x) if(!x %in% installed.packages()){
    install.packages(x, dependencies = T)
  })
  sapply(x, library, character.only=T)
})(x)

```

```{r Uploading file, include=FALSE}
#download.file("https://figshare.com/ndownloader/files/2600235", "Data.dta")
Data <- foreign::read.dta("Data.dta")

```

Objective: Intent to treat: simple, low-intensity, self-management lifestyle intervention (HeLP-her) can prevent weight gain in young to middle-aged women

Primary outcome: weight gain at 1 year. Baseline and within-group differences over time were assessed using paired Student's t tests for continuous variables The effects of the intervention on study outcomes (between group differences) at 1 yr were analysed using linear regression with the variable of interest at 1 yr as the outcome variable, adjusted for baseline values, and obtained robust standard errors to adjust for the clustering effect of town in the regression models (the Huber/White/sandwich estimate of variance)

In the second step, analyses were performed on data that included values imputed using linear regression, multiple imputation with boot strapping.

### Sample size:

The study needed a sample size per group of 196 to demonstrate a 1 kg difference in weight gain with the intervention, considering a standard deviation of 3.5 and power of 80%. However, since the study analysis used 40 location clusters, with a variance inflation factor (VIF) of 1.28 and an intracluster correlation of 0.02, the sample needed was 600, because $VIF=1+(n-1)\rho$ [@ukoumunne2002], which is equivalent to: $n=\frac{(1.28-1)}{0.02}+1$

```{r include=FALSE}
#Sample size
power.t.test(power=.8,
             delta=1,
             sd=3.5,
             type="two.sample")


```

### Summary of the intervention impact

```{r echo=FALSE, message=FALSE, warning=FALSE}
Data <- Data %>% mutate(weight_var= wgt12- weight_base)

p_value <-  Data %>% t.test(.$weight_var~ .$group, data=.) %>% .$p.value %>% round(2)

Data %>%
  group_by(group) %>%
  summarise(n=n(),
            Mean=mean(weight_var, na.rm=T),
            SE=(function(...){
  sqrt(var(..., na.rm=T)/sum(!is.na(...)))
})(weight_var)) %>% 
  ggplot(aes(group, Mean, fill=group))+
  geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE), width=.6)+
  geom_bar(stat="identity")+
   theme(panel.background = element_blank(),
          axis.line = element_line( colour = "black"),
          axis.title=element_text(face="bold"),
          axis.text.x= element_text( face= "bold", colour= "black"),
          axis.text.y= element_text(colour= "black"),
          legend.position="none")+
    xlab(NULL)+
    ylab(paste("Weight variation"))+
  scale_fill_manual(values=c("black", "gray"))+
  ylim(-1, 1)+
  geom_hline(yintercept = 0, linetype="dashed")+
  geom_text(aes(x=group, y=Mean+SE, label= c(NA, paste("* p-value= ", p_value))),
            vjust=12)
```

```{r eval=FALSE, include=FALSE}

#There was no difference in weight gain inside groups:
(sapply(c("Control", "Intervention"), function(x){
  Data %>% filter(group==x) %>% t.test(.$weight_var, data=.)
}))[3,]#Extract p.value

# Map Alternative
# map(.x=c("Control", "Intervention"), .f = function(x){
#  Data %>% filter(group==x) %>% t.test(.$weight_var, data=.) %>% .$p.value
# })



```

### Adjusted analysis

```{r include=FALSE, message=FALSE, warning=FALSE}

#Summary of the model
model <- Data %>% 
  lm(weight_var~ group + BMI_base, data=.)

model %>% summary()

confint(model, level=.95)



#Exploring correlation between variables

Correlations <- Data %>% select(where(is.numeric)) %>% as.matrix() %>% rcorr()


Filtered_correlations <- Correlations$r %>% 
  reshape2::melt() %>% 
  mutate(p.value= Correlations$P %>%
           reshape2::melt() %>%
           select(3) %>%
           round(2) %>%
           unlist(),
         symbol= ifelse(p.value<=0.05, "*", NA)) %>% 
  filter(Var1=="weight_var")








```

The linear model of the intervention impact adjusted by the initial BMI reveals an effect of 0.89kg, with a significant p-value of 0.0184. However, this model does not fit the data adequately. The residual standard error (RSE) is very high (4.157), while the R^2^ is very low, 0.021. One culprit behind this is that having a higher Glycemic Index is a more powerful predictor for weight loss than the intervention.

```{r echo=FALSE, message=FALSE, warning=FALSE}
  Data %>% ggplot(aes(weight_var, GlycemicIndex))+
     theme(panel.background = element_blank(),
          axis.line = element_line( colour = "black"),
          axis.title=element_text(face="bold"),
          axis.text.x= element_text( face= "bold", colour= "black"),
          axis.text.y= element_text(colour= "black"))+
    geom_point(aes(color=group))+
      scale_color_manual(values=c("black", "gray60"))+
    xlab("Weight variation")+
    ylab("Glycemic Index")+
    geom_smooth(method="lm")+
    labs(color="Groups")


```

Therefore, adjusting the linear model by the glycemic index removes the significant effect of the intervention on weight variation and improves the R^2^ from 0.021 to 0.048. While removing the intervention from the model almost does not affect the R^2^ (0.043).

```{r echo=FALSE}

Glycemic_model <- Data %>% lm(weight_var~ GlycemicIndex + group + BMI_base, data=.)
#Glycemic_model %>% summary()

Glycemic_model %>%
  broom::tidy() %>% 
  select(-statistic) %>% 
  kable(caption="Weight variation adjusted by glycemic index") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = T, 
                font_size = 12)


Model_without_intervention <- Data %>% lm(weight_var~ GlycemicIndex + BMI_base, data=.)
#Model_without_intervention %>% summary()

Model_without_intervention %>%
  broom::tidy() %>% 
   select(-statistic) %>% 
  kable(caption="Weight variation adjusted by glycemic index without intervention") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = T, 
                font_size = 12)



```

Therefore, adjusting the linear model by the glycemic index removes the significant effect of the intervention on weight variation and improves the R from 0.02 to 11. While removing the intervention from the model almost does not affect the R (0.043). Thus, the approach of using the glycemic index as a predictor for weight variation renders a better model than the one initially proposed by the study. But more importantly, this data analysis points out that rural women of Australia with higher glycemic index have a greater chance of losing weight, even if they are not more susceptible to the effects of self-management lifestyle intervention.

### References
